---
title: "PCA Analysis (Multiple Filter Scenarios)"
date: "2025-01-22"
output: html_document
---

## Set up your environment and load libraries

```{r}

# If you need to install packages, uncomment the following lines:
# list.of.packages <- c("adegenet", "pegas", "scatterplot3d", "ggplot2", "dplyr", "plotly", "factoextra")
# new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
# if(length(new.packages)) install.packages(new.packages)

library(adegenet)
library(pegas)
library(scatterplot3d)
library(ggplot2)
library(dplyr)
library(plotly)
library(factoextra)


```

## Set working directory and load data

```{r}

# Adjust this path as needed
setwd("/Users/sanzhar/Desktop/pca_0_9")

# 1) Load genotype data
geno_data <- read.table("pca_input.txt", 
                        header=TRUE, 
                        sep="\t",
                        blank.lines.skip=TRUE,
                        stringsAsFactors=FALSE)


# 2) Load sample metadata
metadata <- read.csv("samples_metadata.csv", header=TRUE, stringsAsFactors=FALSE)



```

### PART 1: PCA on the Full, Unfiltered Dataset
## 1.1 Format Genetic Data (Unfiltered)

```{r}

# Make a copy of the full data
geno_data_full <- geno_data
metadata_full  <- metadata

# Extract genotype columns 
geno_cleaned_full <- geno_data_full[, -(1:2)]
# Replace -9 with NA
geno_cleaned_full[geno_cleaned_full == -9] <- NA

# Convert to genind object
obj_full <- df2genind(geno_cleaned_full, ploidy=2, sep="\t", ncode=3)

# Add population group from metadata
pop(obj_full) <- metadata_full$type

# Define colors and shapes (assuming metadata has a 'color' column)
colors_full    <- metadata_full$color
pch_cluster_full <- ifelse(metadata_full$type == "sp", 19, 17)  # 19 for 'sp', 17 for 'cv'


```


## 1.2 Perform PCA (Unfiltered)

```{r}

# Impute missing data
geno_imputed_full <- scaleGen(obj_full, NA.method = "mean")

# PCA with ade4
pca_full <- dudi.pca(geno_imputed_full, scannf=FALSE, scale=TRUE, nf=3)

# Variance explained
variance_full <- 100 * (pca_full$eig / sum(pca_full$eig))
AXE1_full <- round(variance_full[1], 2)
AXE2_full <- round(variance_full[2], 2)
AXE3_full <- round(variance_full[3], 2)

cat("Unfiltered PCA:\n")
cat(paste("PC1:", AXE1_full, "%\n"))
cat(paste("PC2:", AXE2_full, "%\n"))
cat(paste("PC3:", AXE3_full, "%\n"))


```


## 1.3 3D PCA Plot (Unfiltered)

```{r}

# Prepare data for 3D plot
pca_data_full <- data.frame(
  PC1   = pca_full$li[,1],
  PC2   = pca_full$li[,2],
  PC3   = pca_full$li[,3],
  color = colors_full,
  type  = metadata_full$type
)

plot_ly(pca_data_full, x = ~PC1, y = ~PC2, z = ~PC3,
        type = "scatter3d", mode = "markers",
        color = ~color, symbol = ~type,
        marker = list(size = 5)) %>%
  layout(
    title = "3D PCA Plot (Unfiltered)",
    scene = list(
      xaxis = list(title = paste("PC1:", AXE1_full, "%")),
      yaxis = list(title = paste("PC2:", AXE2_full, "%")),
      zaxis = list(title = paste("PC3:", AXE3_full, "%"))
    )
  )


```


## 1.4 2D PCA Plots (Unfiltered)

```{r}

# PC1 vs PC2
plot(pca_full$li[,1], pca_full$li[,2],
     col  = colors_full,
     pch  = pch_cluster_full,
     xlab = paste("PC1:", AXE1_full, "%"),
     ylab = paste("PC2:", AXE2_full, "%"),
     main = "PCA: PC1 vs PC2 (Unfiltered)")

# PC1 vs PC3
plot(pca_full$li[,1], pca_full$li[,3],
     col  = colors_full,
     pch  = pch_cluster_full,
     xlab = paste("PC1:", AXE1_full, "%"),
     ylab = paste("PC3:", AXE3_full, "%"),
     main = "PCA: PC1 vs PC3 (Unfiltered)")

# PC2 vs PC3
plot(pca_full$li[,2], pca_full$li[,3],
     col  = colors_full,
     pch  = pch_cluster_full,
     xlab = paste("PC2:", AXE2_full, "%"),
     ylab = paste("PC3:", AXE3_full, "%"),
     main = "PCA: PC2 vs PC3 (Unfiltered)")



```


## 1.5 Scree Plot (Unfiltered)
```{r}

plot(variance_full, type="b", pch=19,
     xlab="Principal Components",
     ylab="Percentage of Variance Explained",
     main="Scree Plot for PCA (Unfiltered)")

```


## 1.6 Save Outputs (Unfiltered)

```{r}

pdf("PCA_Results_Unfiltered.pdf")
# Plot 2D
plot(pca_full$li[,1], pca_full$li[,2],
     col=colors_full, pch=pch_cluster_full,
     xlab=paste("PC1:", AXE1_full, "%"),
     ylab=paste("PC2:", AXE2_full, "%"),
     main="PC1 vs PC2 (Unfiltered)")
plot(pca_full$li[,1], pca_full$li[,3],
     col=colors_full, pch=pch_cluster_full,
     xlab=paste("PC1:", AXE1_full, "%"),
     ylab=paste("PC3:", AXE3_full, "%"),
     main="PC1 vs PC3 (Unfiltered)")
plot(pca_full$li[,2], pca_full$li[,3],
     col=colors_full, pch=pch_cluster_full,
     xlab=paste("PC2:", AXE2_full, "%"),
     ylab=paste("PC3:", AXE3_full, "%"),
     main="PC2 vs PC3 (Unfiltered)")

# Scree
plot(variance_full, type="b", pch=19,
     xlab="Principal Components",
     ylab="Percentage of Variance Explained",
     main="Scree Plot (Unfiltered)")

dev.off()

```


### PART 2: PCA on Data with “ku_sp” and “bs_sp” Removed
## 2.1 Filter Data

```{r}


# Remove 'ku_sp' and 'bs_sp'
metadata_no_ku_bs <- metadata %>%
  filter(!(site_id %in% c("ku_sp", "bs_sp")))

# Filter geno_data accordingly
geno_data_no_ku_bs <- geno_data %>%
  filter(sample_id %in% metadata_no_ku_bs$sample_id)


```


## 2.2 Format and Convert to genind

```{r}


geno_cleaned_no_ku_bs <- geno_data_no_ku_bs[, -(1:2)]
geno_cleaned_no_ku_bs[geno_cleaned_no_ku_bs == -9] <- NA

obj_no_ku_bs <- df2genind(geno_cleaned_no_ku_bs, ploidy=2, sep="\t", ncode=3)

# Add population group
pop(obj_no_ku_bs) <- metadata_no_ku_bs$type

# Colors and shape
colors_no_ku_bs    <- metadata_no_ku_bs$color
pch_cluster_no_ku_bs <- ifelse(metadata_no_ku_bs$type == "sp", 19, 17)


```

## 2.3 Perform PCA

```{r}

geno_imputed_no_ku_bs <- scaleGen(obj_no_ku_bs, NA.method = "mean")
pca_no_ku_bs          <- dudi.pca(geno_imputed_no_ku_bs, scannf=FALSE, scale=TRUE, nf=3)

variance_no_ku_bs <- 100 * (pca_no_ku_bs$eig / sum(pca_no_ku_bs$eig))
AXE1_no_ku_bs <- round(variance_no_ku_bs[1], 2)
AXE2_no_ku_bs <- round(variance_no_ku_bs[2], 2)
AXE3_no_ku_bs <- round(variance_no_ku_bs[3], 2)

cat("PCA (no ku_sp, no bs_sp):\n")
cat(paste("PC1:", AXE1_no_ku_bs, "%\n"))
cat(paste("PC2:", AXE2_no_ku_bs, "%\n"))
cat(paste("PC3:", AXE3_no_ku_bs, "%\n"))


```

## 2.4 3D PCA Plot

```{r}
pca_data_no_ku_bs <- data.frame(
  PC1   = pca_no_ku_bs$li[,1],
  PC2   = pca_no_ku_bs$li[,2],
  PC3   = pca_no_ku_bs$li[,3],
  color = colors_no_ku_bs,
  type  = metadata_no_ku_bs$type
)

plot_ly(pca_data_no_ku_bs, x = ~PC1, y = ~PC2, z = ~PC3,
        type = "scatter3d", mode = "markers",
        color = ~color, symbol = ~type,
        marker = list(size = 5)) %>%
  layout(
    title = "3D PCA Plot (No ku_sp, No bs_sp)",
    scene = list(
      xaxis = list(title = paste("PC1:", AXE1_no_ku_bs, "%")),
      yaxis = list(title = paste("PC2:", AXE2_no_ku_bs, "%")),
      zaxis = list(title = paste("PC3:", AXE3_no_ku_bs, "%"))
    )
  )


```

## 2.5 2D PCA Plots

```{r}
# PC1 vs PC2
plot(pca_no_ku_bs$li[,1], pca_no_ku_bs$li[,2],
     col = colors_no_ku_bs, pch = pch_cluster_no_ku_bs,
     xlab = paste("PC1:", AXE1_no_ku_bs, "%"),
     ylab = paste("PC2:", AXE2_no_ku_bs, "%"),
     main = "PCA: PC1 vs PC2 (No ku_sp, No bs_sp)")

# PC1 vs PC3
plot(pca_no_ku_bs$li[,1], pca_no_ku_bs$li[,3],
     col = colors_no_ku_bs, pch = pch_cluster_no_ku_bs,
     xlab = paste("PC1:", AXE1_no_ku_bs, "%"),
     ylab = paste("PC3:", AXE3_no_ku_bs, "%"),
     main = "PCA: PC1 vs PC3 (No ku_sp, No bs_sp)")

# PC2 vs PC3
plot(pca_no_ku_bs$li[,2], pca_no_ku_bs$li[,3],
     col = colors_no_ku_bs, pch = pch_cluster_no_ku_bs,
     xlab = paste("PC2:", AXE2_no_ku_bs, "%"),
     ylab = paste("PC3:", AXE3_no_ku_bs, "%"),
     main = "PCA: PC2 vs PC3 (No ku_sp, No bs_sp)")


```

## 2.6 Scree Plot and Saving Outputs

```{r}
# Scree
plot(variance_no_ku_bs, type="b", pch=19,
     xlab="Principal Components",
     ylab="Percentage of Variance Explained",
     main="Scree Plot (No ku_sp, No bs_sp)")

# Save PDF
pdf("PCA_Results_No_ku_bs.pdf")
plot(pca_no_ku_bs$li[,1], pca_no_ku_bs$li[,2],
     col=colors_no_ku_bs, pch=pch_cluster_no_ku_bs,
     xlab=paste("PC1:", AXE1_no_ku_bs, "%"),
     ylab=paste("PC2:", AXE2_no_ku_bs, "%"),
     main="PC1 vs PC2 (No ku_sp, No bs_sp)")
plot(pca_no_ku_bs$li[,1], pca_no_ku_bs$li[,3],
     col=colors_no_ku_bs, pch=pch_cluster_no_ku_bs,
     xlab=paste("PC1:", AXE1_no_ku_bs, "%"),
     ylab=paste("PC3:", AXE3_no_ku_bs, "%"),
     main="PC1 vs PC3 (No ku_sp, No bs_sp)")
plot(pca_no_ku_bs$li[,2], pca_no_ku_bs$li[,3],
     col=colors_no_ku_bs, pch=pch_cluster_no_ku_bs,
     xlab=paste("PC2:", AXE2_no_ku_bs, "%"),
     ylab=paste("PC3:", AXE3_no_ku_bs, "%"),
     main="PC2 vs PC3 (No ku_sp, No bs_sp)")

plot(variance_no_ku_bs, type="b", pch=19,
     xlab="Principal Components", ylab="Variance Explained",
     main="Scree Plot (No ku_sp, No bs_sp)")

dev.off()

# Save updated metadata
write.table(metadata_no_ku_bs, file="filtered_metadata_no_ku_bs.csv",
            sep="\t", row.names=FALSE)


```

### PART 3: PCA on Data with Only “ku_sp” Removed
## 3.1 Filter Data (No ku_sp)

```{r}
metadata_no_ku <- metadata %>%
  filter(site_id != "ku_sp")
geno_data_no_ku <- geno_data %>%
  filter(sample_id %in% metadata_no_ku$sample_id)

```

## 3.2 Format, Convert to genind

```{r}
geno_cleaned_no_ku <- geno_data_no_ku[, -(1:2)]
geno_cleaned_no_ku[geno_cleaned_no_ku == -9] <- NA

obj_no_ku <- df2genind(geno_cleaned_no_ku, ploidy=2, sep="\t", ncode=3)

pop(obj_no_ku) <- metadata_no_ku$type

colors_no_ku    <- metadata_no_ku$color
pch_cluster_no_ku <- ifelse(metadata_no_ku$type == "sp", 19, 17)


```

## 3.3 Perform PCA (No ku_sp)

```{r}
geno_imputed_no_ku <- scaleGen(obj_no_ku, NA.method = "mean")
pca_no_ku          <- dudi.pca(geno_imputed_no_ku, scannf=FALSE, scale=TRUE, nf=3)

variance_no_ku <- 100 * (pca_no_ku$eig / sum(pca_no_ku$eig))
AXE1_no_ku <- round(variance_no_ku[1], 2)
AXE2_no_ku <- round(variance_no_ku[2], 2)
AXE3_no_ku <- round(variance_no_ku[3], 2)


```

## 3.4 3D PCA Plot

```{r}
pca_data_no_ku <- data.frame(
  PC1   = pca_no_ku$li[,1],
  PC2   = pca_no_ku$li[,2],
  PC3   = pca_no_ku$li[,3],
  color = colors_no_ku,
  type  = metadata_no_ku$type
)

plot_ly(pca_data_no_ku, x = ~PC1, y = ~PC2, z = ~PC3,
        type="scatter3d", mode="markers",
        color=~color, symbol=~type,
        marker=list(size=5)) %>%
  layout(
    title = "3D PCA Plot (No ku_sp)",
    scene = list(
      xaxis = list(title = paste("PC1:", AXE1_no_ku, "%")),
      yaxis = list(title = paste("PC2:", AXE2_no_ku, "%")),
      zaxis = list(title = paste("PC3:", AXE3_no_ku, "%"))
    )
  )


```

## 3.5 2D PCA Plots


```{r}
plot(pca_no_ku$li[,1], pca_no_ku$li[,2],
     col=colors_no_ku, pch=pch_cluster_no_ku,
     xlab=paste("PC1:", AXE1_no_ku, "%"),
     ylab=paste("PC2:", AXE2_no_ku, "%"),
     main="PC1 vs PC2 (No ku_sp)")

plot(pca_no_ku$li[,1], pca_no_ku$li[,3],
     col=colors_no_ku, pch=pch_cluster_no_ku,
     xlab=paste("PC1:", AXE1_no_ku, "%"),
     ylab=paste("PC3:", AXE3_no_ku, "%"),
     main="PC1 vs PC3 (No ku_sp)")

plot(pca_no_ku$li[,2], pca_no_ku$li[,3],
     col=colors_no_ku, pch=pch_cluster_no_ku,
     xlab=paste("PC2:", AXE2_no_ku, "%"),
     ylab=paste("PC3:", AXE3_no_ku, "%"),
     main="PC2 vs PC3 (No ku_sp)")


```

## 3.6 Scree Plot & Save Outputs

```{r}
# Scree
plot(variance_no_ku, type="b", pch=19,
     xlab="Principal Components",
     ylab="Percentage of Variance Explained",
     main="Scree Plot (No ku_sp)")

pdf("PCA_Results_No_ku.pdf")
plot(pca_no_ku$li[,1], pca_no_ku$li[,2],
     col=colors_no_ku, pch=pch_cluster_no_ku,
     xlab=paste("PC1:", AXE1_no_ku, "%"),
     ylab=paste("PC2:", AXE2_no_ku, "%"),
     main="PC1 vs PC2 (No ku_sp)")

plot(pca_no_ku$li[,1], pca_no_ku$li[,3],
     col=colors_no_ku, pch=pch_cluster_no_ku,
     xlab=paste("PC1:", AXE1_no_ku, "%"),
     ylab=paste("PC3:", AXE3_no_ku, "%"),
     main="PC1 vs PC3 (No ku_sp)")

plot(pca_no_ku$li[,2], pca_no_ku$li[,3],
     col=colors_no_ku, pch=pch_cluster_no_ku,
     xlab=paste("PC2:", AXE2_no_ku, "%"),
     ylab=paste("PC3:", AXE3_no_ku, "%"),
     main="PC2 vs PC3 (No ku_sp)")

plot(variance_no_ku, type="b", pch=19,
     xlab="Principal Components", ylab="Variance Explained",
     main="Scree Plot (No ku_sp)")

dev.off()

write.table(metadata_no_ku, file="filtered_metadata_no_ku.csv",
            sep="\t", row.names=FALSE)


```

### PART 4: PCA on Data with Only “bs_sp” Removed
## 4.1 Filter Data (No bs_sp)

```{r}

metadata_no_bs <- metadata %>%
  filter(site_id != "bs_sp")

geno_data_no_bs <- geno_data %>%
  filter(sample_id %in% metadata_no_bs$sample_id)

```

## 4.2 Format and Convert to genind

```{r}
geno_cleaned_no_bs <- geno_data_no_bs[, -(1:2)]
geno_cleaned_no_bs[geno_cleaned_no_bs == -9] <- NA

obj_no_bs <- df2genind(geno_cleaned_no_bs, ploidy=2, sep="\t", ncode=3)
pop(obj_no_bs) <- metadata_no_bs$type

colors_no_bs      <- metadata_no_bs$color
pch_cluster_no_bs <- ifelse(metadata_no_bs$type == "sp", 19, 17)


```

## 4.3 Perform PCA (No bs_sp)

```{r}
geno_imputed_no_bs <- scaleGen(obj_no_bs, NA.method = "mean")
pca_no_bs          <- dudi.pca(geno_imputed_no_bs, scannf=FALSE, scale=TRUE, nf=3)

variance_no_bs <- 100 * (pca_no_bs$eig / sum(pca_no_bs$eig))
AXE1_no_bs <- round(variance_no_bs[1], 2)
AXE2_no_bs <- round(variance_no_bs[2], 2)
AXE3_no_bs <- round(variance_no_bs[3], 2)


```

## 4.4 3D PCA Plot

```{r}
pca_data_no_bs <- data.frame(
  PC1   = pca_no_bs$li[,1],
  PC2   = pca_no_bs$li[,2],
  PC3   = pca_no_bs$li[,3],
  color = colors_no_bs,
  type  = metadata_no_bs$type
)

plot_ly(pca_data_no_bs, x = ~PC1, y = ~PC2, z = ~PC3,
        type="scatter3d", mode="markers",
        color=~color, symbol=~type,
        marker=list(size=5)) %>%
  layout(
    title = "3D PCA Plot (No bs_sp)",
    scene = list(
      xaxis = list(title = paste("PC1:", AXE1_no_bs, "%")),
      yaxis = list(title = paste("PC2:", AXE2_no_bs, "%")),
      zaxis = list(title = paste("PC3:", AXE3_no_bs, "%"))
    )
  )


```

## 4.5 2D PCA Plots

```{r}
# PC1 vs PC2
plot(pca_no_bs$li[,1], pca_no_bs$li[,2],
     col=colors_no_bs, pch=pch_cluster_no_bs,
     xlab=paste("PC1:", AXE1_no_bs, "%"),
     ylab=paste("PC2:", AXE2_no_bs, "%"),
     main="PC1 vs PC2 (No bs_sp)")

# PC1 vs PC3
plot(pca_no_bs$li[,1], pca_no_bs$li[,3],
     col=colors_no_bs, pch=pch_cluster_no_bs,
     xlab=paste("PC1:", AXE1_no_bs, "%"),
     ylab=paste("PC3:", AXE3_no_bs, "%"),
     main="PC1 vs PC3 (No bs_sp)")

# PC2 vs PC3
plot(pca_no_bs$li[,2], pca_no_bs$li[,3],
     col=colors_no_bs, pch=pch_cluster_no_bs,
     xlab=paste("PC2:", AXE2_no_bs, "%"),
     ylab=paste("PC3:", AXE3_no_bs, "%"),
     main="PC2 vs PC3 (No bs_sp)")


```

## 4.6 Scree Plot & Save Outputs

```{r}
plot(variance_no_bs, type="b", pch=19,
     xlab="Principal Components",
     ylab="Percentage of Variance Explained",
     main="Scree Plot (No bs_sp)")

pdf("PCA_Results_No_bs.pdf")
plot(pca_no_bs$li[,1], pca_no_bs$li[,2],
     col=colors_no_bs, pch=pch_cluster_no_bs,
     xlab=paste("PC1:", AXE1_no_bs, "%"),
     ylab=paste("PC2:", AXE2_no_bs, "%"),
     main="PC1 vs PC2 (No bs_sp)")

plot(pca_no_bs$li[,1], pca_no_bs$li[,3],
     col=colors_no_bs, pch=pch_cluster_no_bs,
     xlab=paste("PC1:", AXE1_no_bs, "%"),
     ylab=paste("PC3:", AXE3_no_bs, "%"),
     main="PC1 vs PC3 (No bs_sp)")

plot(pca_no_bs$li[,2], pca_no_bs$li[,3],
     col=colors_no_bs, pch=pch_cluster_no_bs,
     xlab=paste("PC2:", AXE2_no_bs, "%"),
     ylab=paste("PC3:", AXE3_no_bs, "%"),
     main="PC2 vs PC3 (No bs_sp)")

plot(variance_no_bs, type="b", pch=19,
     xlab="Principal Components", ylab="Variance Explained",
     main="Scree Plot (No bs_sp)")

dev.off()

write.table(metadata_no_bs, file="filtered_metadata_no_bs.csv",
            sep="\t", row.names=FALSE)


```


